---
title: "TimeseriesAnalysis"
author: "Lawrence Ramsay"
date: "2023-11-14"
output: 
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

devtools::load_all()

options(scipen = 999)
run_hyperparams = F

# istr <- data.table(openxlsx::read.xlsx(paste0(rprojroot::find_rstudio_root_file(), "/data/Instruments.xlsx")))
# 
test_prc_symb <- data.table(
  #"MB-SB-0001", "MB-SB-0002","MB-SB-0003","MB-SB-0004","MB-SB-0006","MB-SB-0009","MB-SB-0010"
  cde = c("close_crd","close_irn"),
  type = c("crude", "ironore"),
  desc = c("Crude Oil","Iron Ore")
) %>% unique() 


#test_prc_symb <- c("MB-IRO-0001","MB-IRO-0002","MB-IRO-0004","MB-IRO-0005","MB-IRO-0006","MB-IRO-0014", "MB-FE-0003")

ex_cols <- c("dte", "Mth", "Currency", "UnitOfMeasure")

#cols = c("Point Forecast", "Lo 80", "Hi 80", "Lo 95", "Hi 95")
cols = c("mn", "lo1", "hi1", "lo2", "hi2", "fcst_obs")

# %>% #unique(fcst_codes$core)%>%
#                                           stringr::str_replace_all("_", "-")

prc_raw <- readr::read_csv(paste0(rprojroot::find_rstudio_root_file() ,"/data/dat_raw.csv"), show_col_types = F) %>%
  data.table() %>%
  .[, .(dte, close_crd, close_irn)] %>%
  copy() %>%
  #set_names(stringr::str_replace_all(names(.), "_", "-")) %>%
  setnames(old = test_prc_symb[1, cde], new = test_prc_symb[1, type]) %>%
  setnames(old = test_prc_symb[2, cde], new = test_prc_symb[2, type]) %>%
  .[, spd := crude / ironore] %>%
  na.omit()

spd_first_val <- prc_raw[1, spd]
crude_first_val <- prc_raw[1, crude]
ironore_first_val <- prc_raw[1, ironore]

prc <- 
  prc_raw %>%
  copy()  %>%
  .[, crude_01 := shift(crude, 1)] %>%
  .[, crude_07 := shift(crude, 7)] %>%
  .[, crude_28 := shift(crude, 28)] %>%
  .[, crude_90 := shift(crude, 90)] %>%
  .[, crude_180 := shift(crude, 180)] %>%
  .[, crude_270 := shift(crude, 270)] %>%
  .[, crude_365 := shift(crude, 365)] %>%
  .[, ironore_01 := shift(ironore, 1)] %>%
  .[, ironore_07 := shift(ironore, 7)] %>%
  .[, ironore_28 := shift(ironore, 28)] %>%
  .[, ironore_90 := shift(ironore, 90)] %>%
  .[, ironore_180 := shift(ironore, 180)] %>%
  .[, ironore_270 := shift(ironore, 270)] %>%
  .[, ironore_365 := shift(ironore, 365)] %>%
  na.omit()

ex_cols <- names(prc)[1:4]

prc_cor <- cor(as.matrix((prc[, !..ex_cols])), prc[, .(crude)])

prc_cor_dt <- 
  data.table(prc_sft = rownames(prc_cor), corr = prc_cor) %>%
  tidyr::separate(prc_sft, sep = "_", into = c("prc", "sft")) %>%
  data.table()# %>%
  #dcast(sft ~ prc, value.var = 'corr.MB_STE_0119') %>%
  #.[order(as.numeric(sft))]

test_points <- 365

dte_all <- data.table(dte = seq.Date(min(prc$dte), max(prc$dte), by = "day"))

prc_all <- merge(dte_all, prc_raw, by = "dte", all.x = T) %>%
  .[, .(dte, crude, ironore, spd)] %>%
  setnafill(type = "locf")

#https://robjhyndman.com/hyndsight/seasonal-periods/

crude_ts <- ts(prc_all[, crude], start=c(year(min(prc_all$dte)),as.numeric(format(prc_all$dte[1], "%j"))),
            frequency = 365)

ironore_ts <- ts(prc_all[, ironore], start=c(year(min(prc_all$dte)),as.numeric(format(prc_all$dte[1], "%j"))),
            frequency = 365)
# 
# spd_ts <- ts(prc[, spd], start=c(year(min(prc$dte)),month(min(prc$dte))), 
#             end=c(year(max(prc$dte)),month(max(prc$dte))), frequency = 365)

crude <- prc[, .(dte, crude)] 
ironore <- prc[, .(dte, ironore)]
spd <- prc[, .(dte, spd)]

```

```{r}

prc[, .N, by = year(dte)]

```


## Overview

This report takes us through forecasting Turkish Rebar and how the effects of seasonality can help the forecasting. This report has quite a lot of experimentation so it meant to be read more like a story. 


It follows the following stages:

1. Visualising the data
2. Viewing the statistical properties 
  - This is to check if the data needs to be transformed to create a better forecast


## Codes

The codes we are using and what they are referring to. I refer to the Turkish price as the crude because that is what we are trying to forecast at the end of the day. The ironore is the price we are comparing to. This can be anything we think could be similar.  

```{r}

reactable::reactable(test_prc_symb)

```


## Visualisation

First off start by viewing the data. The spread might also be a useful way of seeing and predicting the data as it can remove some ofthe volatility. In this case the spread is done as a ratio to normalise the data. 

```{r, fig.width=9}

p1 <- ggplot(prc_raw) +
  geom_line(aes(dte, crude)) +
  labs(title = "Raw Timeseries for Brent Crude and Iron Ore", x = NULL, y = "Crude Price ($/bl)")

p2 <- ggplot(prc_raw) +
  geom_line(aes(dte, ironore)) +
  labs(x = NULL, y = "Iron Ore Price ($/tonne)")

p3 <- prc %>%
  na.omit() %>%
  ggplot() +
  geom_line(aes(dte, spd)) +
  labs(x = "Date", y = "Spread (ratio)")

p1 / p2 / p3
```

When we scale to show the prices and spread together with smoothing we can see the spread is fairly consistent until after the big spike in prices where one diverges from the other significantly. This means the spread is not immune to the forces of the other prices so we will need to deal with high voltility either way. 

```{r, fig.height=6, fig.width=9}

prc_scale <- 
  data.table(dte = prc[, dte],
             Spread = scale(prc[, spd], center = F)[, 1],
             Crude = scale(prc[["crude"]], center = F)[, 1],
             `Iron Ore` = scale(prc[["ironore"]], center = F)[, 1]) %>%
  melt(id.vars = "dte", variable.name = "Price") 

ggplot(prc_scale) +
  geom_smooth(aes(dte, value, color = Price), span = 0.05, method = "loess") +
  labs(title = "Scaled Timeseries of Crude, Brent and Spread",
       y = "Scaled Price (between 0 and 1)", x = "Date")
  

```

### Seasonality

The price itself shows a complex seasonality but it's fairly weak. The randomness is much higher than the seasonality. It means the forecast using the past price patterns alone. There is a good candidate in the 365 area, so what was happening a the same time last year looks to have some indication. 

More importantly this yearly cycle looks deterministic which should help with prediction. 

More info on seasonality types here:
https://towardsdatascience.com/3-types-of-seasonality-and-how-to-detect-them-4e03f548d167

When we talk about 28, 90 that is the seasonal period that is averaged together to create the seasonal pattern. As the series is daily, the frequency per year is 365 then we divide by 28 to approximate a month. It dosnt' have to be exact as there are many influences on seasonal patterns. 

More info here:
https://otexts.com/fpp2/classical-decomposition.html

```{r, fig.height=7, fig.width=12}
# st_ts <- ts(prc[,crude], start=c(year(min(prc$dte)),month(min(prc$dte))), 
#             end=c(year(max(prc$dte)),month(max(prc$dte))),frequency = c(365))

plt_ts_one <- forecast::autoplot(decompose(crude_ts)) + 
  labs(title = "Crude oil decomposition with single and multiple seasonalities")

crude_msts <- forecast::msts(prc[,crude], start=c(year(min(prc$dte)),month(min(prc$dte))), 
                          end=c(year(max(prc$dte)),month(max(prc$dte))),seasonal.periods = c(7, 28, 90, 365), ts.frequency = 365)

plt_ts_mult <- forecast::mstl(crude_msts, s.window = c(7)) %>% autoplot()


plt_ts_one | plt_ts_mult


```

```{r, fig.height=7, fig.width=12}
# st_ts <- ts(prc[,crude], start=c(year(min(prc$dte)),month(min(prc$dte))), 
#             end=c(year(max(prc$dte)),month(max(prc$dte))),frequency = c(365))

plt_ts_one <- forecast::autoplot(decompose(ironore_ts))  + 
  labs(title = "Iron ore decomposition with single and multiple seasonalities")

ironore_msts <- forecast::msts(prc[,ironore], start=c(year(min(prc$dte)),month(min(prc$dte))), 
                          end=c(year(max(prc$dte)),month(max(prc$dte))),seasonal.periods = c(7, 28, 90, 365), ts.frequency = 365)

plt_ts_mult <- forecast::mstl(ironore_msts, s.window = c(7)) %>% autoplot()


plt_ts_one | plt_ts_mult


```

## Statistical properties

Broadly speaking the steps are taken from this section:
https://otexts.com/fpp2/arima-r.html

### ACF and PACF

```{r}

(forecast::ggAcf(crude[, crude], main = "Crude Oil")  / 
  forecast::ggAcf(ironore[[2]], main = "Iron Ore") / 
  forecast::ggAcf(spd[[2]], main = "Spread")) |

(forecast::ggPacf(crude[[2]], main = "Crude Oil") / 
   forecast::ggPacf(ironore[[2]], main = "Iron Ore") / 
   forecast::ggPacf(spd[[2]], main = "Spread") )
```

### Checking Normality

Checking for normality isn't essential as you can still get predictions but it can help get better results as most of the basic models 

Looking at this data, we can see the data is not normally distributed and skewed to the left. This is very common for financial data and to be expected. 

The spread is the best so it might be better for making predictions. 

```{r}

prc %>%
 .[, .(dte, Spread = spd, Crude = crude, `Iron Ore` = ironore)] %>%
  melt(id.vars = "dte") %>%
ggplot() +
  #geom_histogram(aes(value)) +
  geom_density(aes(value)) +
  facet_wrap(vars(variable), scales = "free")  +
  labs(title = "Density plot of each price", x = "Price", y = "Density")

```

This is just a p-value test of the above, but it is saying the same thing. Anything < 0.01 is not normally distributed.

```{r}

shapiro.test(crude[[2]])
shapiro.test(ironore[[2]])
shapiro.test(spd[[2]])

```

### Checking Variance

Another problem with financial timeseries is changing variance (volatility) of the data. If variance is constantly changing it can affect predictions. Again, predictions can still be made, but it might help accuracy to reduce the variance before making predictions.

Below is a plot of standard deviation by month. This is to find any changing variance

```{r}
prc[, .(Spread = sd(spd),
        Crude = sd(crude),
        `Iron Ore` = sd(ironore)), by = floor_date(dte, "month")] %>%
  setnames(old = "floor_date", new = "Mth") %>%
  melt("Mth") %>%
  ggplot() +
  geom_line(aes(Mth, value)) +
  facet_grid(rows = vars(variable), scales = "free_y") +
  labs(title = "Standard Deviation of Prices Over Time", x = "Date", y = "Standard Deviation")
```

This plot uses the changepoint package which uses a different measure to find the change in variance. 

```{r, fig.height=8}

par(mfrow = c(3, 1))

crude_var <- changepoint::cpt.var(prc[, crude])

plot(crude_var, main = "Crude Variance Changepoint Analysis", ylab = "Price ($/bl)")

ironore_var <- changepoint::cpt.var(prc[, ironore])

plot(ironore_var, main = "Iron Ore", ylab = "Price ($/tonne)")

spd_var <- changepoint::cpt.var(prc[, spd], mu = mean(prc[, spd]), know.mean = T)

plot(spd_var, main = "Spread", ylab = "Price (ratio)")

```

```{r, fig.height=8}

par(mfrow = c(3, 1))

crude_var <- changepoint::cpt.mean(prc[, crude])

plot(crude_var, main = "Crude Mean Changepoint Analysis", ylab = "Price ($/bl)")

ironore_var <- changepoint::cpt.mean(prc[, ironore])

plot(ironore_var, main = "Ironore", ylab = "Price ($/bl)")

spd_var <- changepoint::cpt.mean(prc[, spd])

plot(spd_var, main = "Spread", ylab = "Price ($/bl)")

```

In conclusion the variance does change. Again, this is normal for financial data which tend to go though periods where it is stable and more volatile. 

## Explanatory Feature

### Correlation

It was mentioned that 0417 can be used to explain the activity of 0119. The thing that is worth bearing in mind here is they could be correlated this doesn't mean one is good a predicting the other. Most of the forecast heavy lifting is done through analysing the price itself and features that are predictive. Things like the seasonality. 

This shows there is a heavy correlation (to be expected), but that the means are significantly different from each other (t-test)

In the 400 bounds of 417 (US) up to 0600 bounds of 0119  (Turkey), the correlation is tight, meaning they move together very tightly. Then above this they tend to break apart. This is also what we swa 

```{r, fig.height=7, fig.width=8}

ggstatsplot::ggscatterstats(data = prc, x = crude, y = ironore)

```


### Time Shifting

As the prices are shifted against each other overtime this looks at the degradation of the correlation. What it essentially means is the US price is better at predicting the price over the longer term. This means it could be useful for forecasting, more useful than the price itself at least which would be considered the  baseline. The differences are pretty marginal however. 

```{r}
prc_cor_dt %>%
.[, .(Price = prc, Shift = sft, Correlation = corr.crude)] %>%
.[Price == "crude", Price := "Crude"] %>%
.[Price == "ironore", Price := "Iron Ore"] %>%
ggplot() +
  geom_line(aes(as.numeric(Shift), Correlation, color = Price), linewidth = 1) +
  labs(title = "Auto Correlation from 1 to 365", x = "Time Shift", y = "Pearson Correlation")


```

## Inital Modelling

At this point you can put the values straight into auto.arima and see how it performs. This is because the forecast package can do most of the work for you in the modelling process or at least find a best guess. 

I take off the last 365 points (test set) and then test the forecast against this test set. The lower the RMSE (error rate) the better. Bear in mind it is scale dependent so the error rate cannot be compared across the crude and spread models. 

I have added the residuals (difference between observed and predicted) to see how the model performs. For the timeline we want no discernible pattern. 

The ACF of residuals looks for auto correlations (patterns from previous values). There is a high auto-correlation 5, 10 and 15 days previous to each time point. This is a problem since the model hasnt removed it in it's process but might notbe the end of the world. 

Finally the historgram is the distribution of the residuals, we essentially want it to center around 0 and be even on both sides. 

```{r, fig.width=10}

train_test_split_forecast <- function(y_dt, test_points = 365, level = c(80, 95), auto = T, order = c(0,0,0), seasonal = c(0,0,0),
                                      xreg_train = NULL, xreg_test = NULL) {
  
    train <- y_dt[1:(nrow(y_dt) - test_points)][[2]]
    test <- y_dt[(nrow(y_dt) - test_points + 1):nrow(y_dt)][[2]]
    train_dtes <- y_dt[1:(nrow(y_dt) - test_points), dte]
    test_dtes <- y_dt[(nrow(y_dt) - test_points):nrow(y_dt), dte][-1]
    
    if (!is.null(xreg_train)) {
      xreg_train <- matrix(xreg_train[1:(nrow(xreg_train) - test_points)][[2]])
      
      if (auto) {
        mdl <- forecast::auto.arima(train, xreg = xreg_train)
      } else {
        mdl <- forecast::Arima(train, order = order, seasonal = seasonal, xreg = xreg_train)
      }
      
      fcst <- forecast::forecast(mdl, h = test_points, level = level, xreg = xreg_test)
      
    } else {
      
      if (auto) {
        mdl <- forecast::auto.arima(train)
      } else {
        mdl <- forecast::Arima(train, order = order, seasonal = seasonal)
      }
      
      fcst <- forecast::forecast(mdl, h = test_points, level = level)
      
    }
    
    fcst_dt <- data.table(dte = test_dtes, mn = fcst$mean, lo1 = fcst$lower[,1], lo2 = fcst$lower[,2],
                          hi1 = fcst$upper[,1], hi2 = fcst$upper[,2], fcst_obs = test)
    
    train_dt <- data.table(dte = train_dtes, y_val = train)
    
    return(list(fcst_dt, fcst, train_dt))
  
}

train_test_split_forecast_ts <- function(y_dt, test_points = 365, 
                                         level = c(80, 95), auto = T, order = c(0,0,0), seasonal = c(0,0,0),
                                      xreg_train = NULL, xreg_test = NULL) {

    train_dtes <- y_dt[1:(nrow(y_dt) - test_points), dte]
    test_dtes <- y_dt[(nrow(y_dt) - test_points):nrow(y_dt), dte][-1]
  
    train <- ts(y_dt[dte %in% train_dtes][[2]], start=c(year(train_dtes[1]),as.numeric(format(train_dtes[1], "%j"))),
              frequency = 365)
  
    test <- ts(y_dt[dte %in% test_dtes][[2]], start=c(year(test_dtes[1]),as.numeric(format(test_dtes[1], "%j"))),
                frequency = 365)

    if (!is.null(xreg_train)) {
      xreg_train <- matrix(xreg_train[1:(nrow(xreg_train) - test_points)][[2]])
      
      if (auto) {
        mdl <- forecast::auto.arima(train, xreg = xreg_train)
      } else {
        mdl <- forecast::Arima(train, order = order, seasonal = seasonal, xreg = xreg_train)
      }
      
      fcst <- forecast::forecast(mdl, h = test_points, level = level, xreg = xreg_test)
      
    } else {
      
      if (auto) {
        mdl <- forecast::auto.arima(train)
      } else {
        mdl <- forecast::Arima(train, order = order, seasonal = seasonal)
      }
      
      fcst <- forecast::forecast(mdl, h = test_points, level = level)
      
    }
    
    fcst <- forecast::forecast(mdl, h = test_points, level = level, xreg = xreg_test)
    
    fcst_dt <- data.table(dte = test_dtes, mn = fcst$mean, lo1 = fcst$lower[,1], lo2 = fcst$lower[,2],
                          hi1 = fcst$upper[,1], hi2 = fcst$upper[,2], fcst_obs = test)
    
    train_dt <- data.table(dte = train_dtes, y_val = train)
    
    return(list(fcst_dt, fcst, train_dt))
  
}

plot_fcst_resid <- function(y_dt, fcst_dt, fcst, train_dt, vis_start_date = lubridate::as_date("1900-01-01")) {
  
    acc <- forecast::accuracy(as.numeric(y_dt[dte %chin% fcst_dt[, dte]][[2]]), as.numeric(fcst_dt$mn))
    
    y_dt_vis <- y_dt[dte >= vis_start_date]
    fcst_dt_vis <- fcst_dt[dte >= vis_start_date]
    train_dt_vis <- train_dt[dte >= vis_start_date]
  
    p1 <- ggplot() +
    geom_line(aes(y_dt_vis$dte, y_dt_vis[[2]])) +
    geom_line(aes(fcst_dt_vis$dte, fcst_dt_vis$mn), linetype = "dashed", color = "darkred") +
    geom_ribbon(aes(fcst_dt_vis$dte, ymin = fcst_dt_vis$lo1, ymax = fcst_dt_vis$hi1), alpha = 0.2, fill = "red") +
    geom_ribbon(aes(fcst_dt_vis$dte, ymin = fcst_dt_vis$lo2, ymax = fcst_dt_vis$hi2), alpha = 0.4, fill = "red") +
        geom_text() +
        annotate(
          "text", label = paste0("RMSE: ", round(acc[2],2)),
          x = lubridate::as_date(max(y_dt_vis[(length(y_dt_vis[, dte])-200), dte])), 
          y = min(y_dt_vis[[2]]), size = 4, colour = "darkred"
        ) +
    labs(title = paste0("Forecast for ", names(y_dt_vis)[2], " ", fcst$method))
  
  p2 <- ggplot() +
        geom_line(aes(x = train_dt$dte, y = as.numeric(fcst$residuals))) +
    labs(y = "Train Residuals")
  
  p3 <- forecast::ggAcf(as.numeric(fcst$residuals)) +
    labs(title = "ACF of Residuals")
  
  p4 <- ggplot() +
    geom_histogram(aes(as.numeric(fcst$residuals)), bins = 40) +
    labs(title = "Histogram of Residuals", x = "Spread")
  
    p1 / p2 / (p3 | p4)
 
}

fcst_list <- train_test_split_forecast(y_dt = crude, test_points = test_points, xreg_train = NULL)

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_list[[1]], fcst = fcst_list[[2]], 
                train_dt = fcst_list[[3]], vis_start_date = lubridate::as_date("2020-01-01"))

```

When testing the spread with the inversion, it doesn't seem to help much. 

```{r, fig.width=10}

fcst_list_spd <- train_test_split_forecast(y_dt = spd, test_points = test_points)

fcst_list_ironore <- train_test_split_forecast(y_dt = ironore, test_points = test_points)

fcst_dt_invert <- fcst_list_spd[[1]] %>%
  copy()

for (col in 2:ncol(fcst_list_ironore[[1]])) {
  
  fcst_dt_invert[[col]] <- fcst_dt_invert[[col]] * fcst_list_ironore[[1]][[col]] 
  
}

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_dt_invert, fcst = fcst_list_spd[[2]], 
                train_dt = fcst_list_spd[[3]], vis_start_date = lubridate::as_date("2020-01-01"))

```

## Transformation

To try to deal with the changing variance and normality that was shown to be the case above we are going to transform the data using a Box-cox.    

```{r, fig.width=12}
#https://www.css.cornell.edu/faculty/dgr2/_static/files/R_html/Transformations.html

crude_comp_lm <- lm(crude$crude ~ ironore$ironore)

summary(crude_comp_lm)
```


```{r, fig.width=12}
par(mfrow=c(2,2))
plot(crude_comp_lm, main = "crude_comp")
```


```{r}
lambda = 0.5

lambda_crude <- forecast::BoxCox.lambda(crude)
crude_bxcx_auto <- data.table(dte = crude[, dte], crude = forecast::BoxCox(crude[[2]], lambda = lambda_crude))
crude_bxcx_manu <- data.table(dte = crude[, dte], crude = forecast::BoxCox(crude[[2]], lambda = lambda))

lambda_ironore <- forecast::BoxCox.lambda(ironore)
ironore_bxcx_auto <- data.table(dte = ironore[, dte], ironore = forecast::BoxCox(ironore[[2]], lambda = lambda_ironore))
ironore_bxcx_manu <- data.table(dte = ironore[, dte], ironore = forecast::BoxCox(ironore[[2]], lambda = lambda))

lambda_spd <- forecast::BoxCox.lambda(spd)
spd_bxcx_auto <- data.table(dte = spd[, dte], spd = forecast::BoxCox(spd[[2]], lambda = lambda_spd))
spd_bxcx_manu <- data.table(dte = spd[, dte], spd = forecast::BoxCox(spd[[2]], lambda = lambda))

# crude_diff <- diff(crude, lag = 1)
# ironore_diff <- diff(ironore, lag = 1)
# spd_diff <- diff(spd, lag = 1)

par(mfrow = c(3, 3))
plot(density(crude[[2]]))
plot(density(ironore[[2]]))
plot(density(spd[[2]]))

plot(density(crude_bxcx_auto[[2]]))
plot(density(ironore_bxcx_auto[[2]]))
plot(density(spd_bxcx_auto[[2]]))

plot(density(crude_bxcx_manu[[2]]))
plot(density(ironore_bxcx_manu[[2]]))
plot(density(spd_bxcx_manu[[2]]))



```

```{r}
#spd_bxcx_inv <- forecast::InvBoxCox(spd_bxcx, lambda = 0.5)
```

Automatic lambda is working the best here so this will be used for the BoxCox

```{r, fig.height=6, fig.width=10}
par(mfrow = c(3, 3))
qqnorm(crude[[2]], main = "crude")
qqline(crude[[2]]) 
qqnorm(ironore[[2]], main = "ironore")
qqline(ironore[[2]]) 
qqnorm(spd[[2]], main = "spd")
qqline(spd[[2]]) 

qqnorm(crude_bxcx_auto[[2]], main = "crude boxcox auto")
qqline(crude_bxcx_auto[[2]]) 
qqnorm(ironore_bxcx_auto[[2]], main = "ironore boxcox auto")
qqline(ironore_bxcx_auto[[2]]) 
qqnorm(spd_bxcx_auto[[2]], main = "spd boxcox auto") 
qqline(spd_bxcx_auto[[2]]) 

qqnorm(crude_bxcx_manu[[2]], main = "crude boxcox manual")
qqline(crude_bxcx_manu[[2]]) 
qqnorm(ironore_bxcx_manu[[2]], main = "ironore boxcox manual")
qqline(ironore_bxcx_manu[[2]]) 
qqnorm(spd_bxcx_manu[[2]], main = "spd boxcox manual") 
qqline(spd_bxcx_manu[[2]]) 

crude_bxcx <- crude_bxcx_auto
ironore_bxcx <- ironore_bxcx_auto
spd_bxcx <- spd_bxcx_auto

```

```{r}

# fcst_arima <- function(y, auto = T, order = c(0,0,0), seasonal = c(0,0,0)) {
#   
#   if (auto) {
#     
#     mdl <- forecast::auto.arima(y)
#   
#   } else {
#     
#     mdl <- forecast::Arima(y, order = order, seasonal = seasonal)
#     
#   }
# 
#   fcst <- forecast::forecast(mdl, h = test_points)
#   
#   acc <- forecast::accuracy(y[(nrow(prc) - test_points):nrow(prc)], fcst$mean)
#   
#   print(mdl)
#   
#   print(acc)
#   
#   print(forecast::checkresiduals(mdl))
#   
#   print(forecast::autoplot(fcst))
#   
# }
# 
# 
# fcst_arima(crude)

```

## Auto with Transformation

### Transformed crude Forecast

```{r, fig.width=10}

fcst_list <- train_test_split_forecast(y_dt = crude_bxcx_auto)

fcst_dt_invert <- fcst_list[[1]] %>%
  copy()%>%
  .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda_crude)}), .SDcols = cols]

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_dt_invert, fcst = fcst_list[[2]], train_dt = fcst_list[[3]])


```

```{r, fig.width=10}

fcst_list <- train_test_split_forecast(y_dt = crude_bxcx_manu)

fcst_dt_invert <- fcst_list[[1]] %>%
  copy()%>%
  .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda)}), .SDcols = cols]

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_dt_invert, fcst = fcst_list[[2]], train_dt = fcst_list[[3]])


```

### Transformed Spread Forecast

```{r, fig.width=10}

fcst_list_spd <- train_test_split_forecast(y_dt = spd_bxcx_auto)

fcst_list_ironore <- train_test_split_forecast(y_dt = ironore_bxcx_auto, test_points = test_points)

fcst_dt_invert <- fcst_list_spd[[1]] %>%
  copy() %>% 
  .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda_spd)}), .SDcols = cols]

for (col in 2:ncol(fcst_list_ironore[[1]])) {
  
  fcst_dt_invert[[col]] <- fcst_dt_invert[[col]] * fcst_list_ironore[[1]][[col]] 
  
}

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_dt_invert, fcst = fcst_list_spd[[2]], train_dt = fcst_list_spd[[3]])


```

```{r, fig.width=10}

fcst_list_spd <- train_test_split_forecast(y_dt = spd_bxcx_manu)

fcst_list_ironore <- train_test_split_forecast(y_dt = ironore_bxcx_auto, test_points = test_points)

fcst_dt_invert <- fcst_list_spd[[1]] %>%
  copy() %>% 
  .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda_spd)}), .SDcols = cols]

for (col in 2:ncol(fcst_list_ironore[[1]])) {
  
  fcst_dt_invert[[col]] <- fcst_dt_invert[[col]] * fcst_list_ironore[[1]][[col]] 
  
}

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_dt_invert, fcst = fcst_list_spd[[2]], train_dt = fcst_list_spd[[3]])


```

In conclusion, the BoxCox transformation doesn't make much difference to the prediction capability. 

## Auto With Diff

It was mentioned earlier that there are auto correlations in the data, it might help to differnce the data first. This just subtracting one value from another to make the data stationary (without trends). 

```{r}
forecast::ggAcf(crude_bxcx[[2]]) / forecast::ggAcf(ironore_bxcx[[2]]) / forecast::ggAcf(spd_bxcx[[2]]) 
forecast::ggPacf(crude_bxcx[[2]]) / forecast::ggPacf(ironore_bxcx[[2]]) / forecast::ggPacf(spd_bxcx[[2]]) 
```

So by far the strongest auto correlation is 1, so we will diff 1. 

If interested in what these plots mean more info here:

https://towardsdatascience.com/interpreting-acf-and-pacf-plots-for-time-series-forecasting-af0d6db4061c

```{r}

crude_bxcx_diff1 <- data.table(dte = crude_bxcx[, dte][-1], crude = diff(crude_bxcx[[2]], lag = 1))
ironore_bxcx_diff1 <- data.table(dte = ironore_bxcx[, dte][-1], ironore = diff(ironore_bxcx[[2]], lag = 1))
spd_bxcx_diff1 <- data.table(dte = spd_bxcx[, dte][-1], spd = diff(spd_bxcx[[2]], lag = 1))

forecast::ggAcf(crude_bxcx_diff1[[2]]) / forecast::ggAcf(ironore_bxcx_diff1[[2]]) / forecast::ggAcf(spd_bxcx_diff1[[2]]) 
forecast::ggPacf(crude_bxcx_diff1[[2]]) / forecast::ggPacf(ironore_bxcx_diff1[[2]]) / forecast::ggPacf(spd_bxcx_diff1[[2]]) 

```

When the data has been diffed to the first value it looks better but there are still heavy auto-correaltions to the 5th day for the crude and spread (to be expected since spread is with crude). This is because the value change weekly but the frequency is daily. This is a problem with the data. 

```{r}
# spd_bxcx <- forecast::BoxCox(spd, lambda = 0.5)
# spd_bxcx_inv <- forecast::InvBoxCox(spd_bxcx, lambda = 0.5)
# 
# spd[1:10]
# spd_bxcx[1:10]
# spd_bxcx_inv[1:10]
```

```{r, fig.width=10}

# crude_bxcx_first_val <- crude_bxcx[1, crude]
# 
# fcst_list <- train_test_split_forecast(y_dt = crude_bxcx_diff1)
# 
# crude_bxcx_inv <- data.table(dte = crude_bxcx[, dte], crude = diffinv(crude_bxcx_diff1[[2]], xi = crude_bxcx_first_val))
# 
# crude_bxcx_fcst_first_val <- crude_bxcx_inv[dte == fcst_list[[1]][1, dte], crude]                
# 
# fcst_dt_invert <- fcst_list[[1]] %>%
#   copy() %>%
#   .[, mn := diffinv(mn, xi = crude_bxcx_fcst_first_val, lag = 1)[-1]] %>%
#   .[, lo1 := mn + lo1] %>%
#   .[, lo2 := mn + lo2] %>%
#   .[, hi1 := mn + hi1] %>%
#   .[, hi2 := mn + hi2] %>%
#   .[, fcst_obs := crude_bxcx_fcst_first_val + fcst_obs] %>%
#   .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda_crude)}), .SDcols = cols]
#   
# plot_fcst_resid(y_dt = crude, fcst_dt = fcst_dt_invert, fcst = fcst_list[[2]], train_dt = fcst_list[[3]])

```

```{r}

crude_bxcx_diff5 <- data.table(dte = crude_bxcx[6:nrow(crude_bxcx), dte], crude = diff(crude_bxcx[[2]], lag = 5))
ironore_bxcx_diff5 <- data.table(dte = ironore_bxcx[6:nrow(crude_bxcx), dte], crude = diff(ironore_bxcx[[2]], lag = 5))
spd_bxcx_diff5 <- data.table(dte = spd_bxcx[6:nrow(crude_bxcx), dte], crude = diff(spd_bxcx[[2]], lag = 5))

forecast::ggAcf(crude_bxcx_diff5[[2]]) / forecast::ggAcf(ironore_bxcx_diff5[[2]]) / forecast::ggAcf(spd_bxcx_diff5[[2]])
forecast::ggPacf(crude_bxcx_diff5[[2]]) / forecast::ggPacf(ironore_bxcx_diff5[[2]]) / forecast::ggPacf(spd_bxcx_diff5[[2]])

```

We can double diff as well. In this case however, it doesnt seem to help. 

```{r}

crude_bxcx_diff15 <- data.table(dte = crude_bxcx[6:nrow(crude_bxcx_diff1), dte], crude = diff(crude_bxcx_diff1[[2]], lag = 5))
ironore_bxcx_diff15 <- data.table(dte = ironore_bxcx[6:nrow(crude_bxcx_diff1), dte], crude = diff(ironore_bxcx_diff1[[2]], lag = 5))
spd_bxcx_diff15 <- data.table(dte = spd_bxcx[6:nrow(crude_bxcx_diff1), dte], crude = diff(spd_bxcx_diff1[[2]], lag = 5))

forecast::ggAcf(crude_bxcx_diff15[[2]]) / forecast::ggAcf(ironore_bxcx_diff15[[2]]) / forecast::ggAcf(spd_bxcx_diff15[[2]])
forecast::ggPacf(crude_bxcx_diff15[[2]]) / forecast::ggPacf(ironore_bxcx_diff15[[2]]) / forecast::ggPacf(spd_bxcx_diff15[[2]])

```

### Diff Modelling

```{r, fig.width=10}

fcst_bxcx_diff <- function(y_dt, y_dt_diff, y_dt_bxcx, lambda, lag_order = 1, diff_invert_hilo = FALSE, hilo_simulate_adjust = 1,
                           auto = T, order = c(0,0,0), seasonal = c(0,0,0), vis_start_date = lubridate::as_date("1990-01-01"),
                           test_points = 365) {
  
  fcst_list <- train_test_split_forecast(y_dt = y_dt_diff, auto = auto, order = order, 
                                         seasonal = seasonal, test_points = test_points)

  crude_bxcx_first_val <- y_dt_bxcx[1:lag_order][[2]]
  
  crude_bxcx_inv <- data.table(dte = y_dt_bxcx[, dte], crude = diffinv(y_dt_diff[[2]],                                                                           xi = crude_bxcx_first_val, lag = lag_order))
  
  crude_bxcx_fcst_first_val <- crude_bxcx_inv[dte %in% fcst_list[[1]]$dte][1:lag_order][[2]]
  
  if (diff_invert_hilo) {
    
    fcst_dt_invert <- fcst_list[[1]] %>%
      copy() %>%
      .[, mn := diffinv(mn, xi = crude_bxcx_fcst_first_val, lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
      .[, lo1 := diffinv(lo1, xi = crude_bxcx_fcst_first_val, lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
      .[, lo2 := diffinv(lo2, xi = crude_bxcx_fcst_first_val, lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
      .[, hi1 := diffinv(hi1, xi = crude_bxcx_fcst_first_val, lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
      .[, hi2 := diffinv(hi2, xi = crude_bxcx_fcst_first_val, lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
      .[, fcst_obs := diffinv(fcst_obs, xi = crude_bxcx_fcst_first_val, 
                              lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
      .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda)}), .SDcols = cols]

  } else {
      fcst_dt_invert <- fcst_list[[1]] %>%
        copy() %>%
        .[, mn := diffinv(mn, xi = crude_bxcx_fcst_first_val, lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
        .[, lo1 := min(mn) + min(lo1) * (hilo_simulate_adjust*(.I/max(.I))+((.I/max(.I))^2))] %>%
        .[, lo2 := min(mn) + min(lo2) * (hilo_simulate_adjust*(.I/max(.I))+((.I/max(.I))^2))] %>%
        .[, hi1 := max(mn) + max(hi1) * (hilo_simulate_adjust*(.I/max(.I))+((.I/max(.I))^2))] %>%
        .[, hi2 := max(mn) + max(hi2) * (hilo_simulate_adjust*(.I/max(.I))+((.I/max(.I))^2))] %>%
        .[, fcst_obs := diffinv(fcst_obs, xi = crude_bxcx_fcst_first_val, 
                                lag = lag_order)[(lag_order + 1):(test_points + lag_order)]] %>%
        .[, (cols) := lapply(.SD, function(x){forecast::InvBoxCox(x, lambda = lambda)}), .SDcols = cols]
  }
  
  plot_fcst_resid(y_dt = y_dt, fcst_dt = fcst_dt_invert, fcst = fcst_list[[2]], 
                  train_dt = fcst_list[[3]], vis_start_date  = vis_start_date)
}

fcst_bxcx_diff(y_dt = crude, y_dt_diff = crude_bxcx_diff1, y_dt_bxcx = crude_bxcx, lag_order = 1, diff_invert_hilo = F,
               lambda = lambda_crude)

```

So the diff1 model is worse

```{r, fig.width=10}

fcst_bxcx_diff(y_dt = crude, y_dt_diff = crude_bxcx_diff5, y_dt_bxcx = crude_bxcx, 
               lag_order = 5, diff_invert_hilo = F, lambda = lambda_crude)

```

The diff5 model however is better than the automatically created model. 

## Other Frequencies

## Crude With Regressor

As mentioned earlier the regressor looks to be more predictive than the price itself, Therefore we are going to project the the comparison price and regress it onto the crude to see if it gives a better prediction. 
 
```{r}

fcst_list_crude <- train_test_split_forecast(y_dt = crude, test_points = test_points)

fcst_list_crude_reg <- train_test_split_forecast(y_dt = crude, test_points = test_points, 
                                                  xreg_train = ironore,
                                                  xreg_test = fcst_list_crude[[1]]$mn
                                                  )

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_list_crude_reg[[1]], fcst = fcst_list_crude_reg[[2]], 
                train_dt = fcst_list_crude_reg[[3]], vis_start_date = lubridate::as_date("2020-01-01"))

```

```{r}

row_crude <- nrow(crude)

crude_diff365 <- data.table(dte = crude[366:row_crude, dte], crude = diff(crude[[2]], lag = 365))

fcst_list_crude <- train_test_split_forecast(y_dt = crude_diff365, test_points = test_points)

fcst_list_crude_reg <- train_test_split_forecast(y_dt = crude[(row_crude-nrow(crude_diff365)+1):row_crude], test_points = test_points, 
                                                  xreg_train = crude_diff365,
                                                  xreg_test = fcst_list_crude[[1]]$mn
                                                  )

plot_fcst_resid(y_dt = crude, fcst_dt = fcst_list_crude_reg[[1]], fcst = fcst_list_crude_reg[[2]], 
                train_dt = fcst_list_crude_reg[[3]], vis_start_date = lubridate::as_date("2020-01-01"))

```

It looks like the models are struggling to find the patterns within the data so we are going to need to force it in slightly. Below is the forecast using the seasonally decomposed forecast of the comparison as a regressor. This means it wont have ans influence on the modelling process but the main aspects of the crude forecast will take precedence. 

```{r}
row_crude <- nrow(crude)

crude_seasonal <- decompose(crude_ts)$seasonal

crude_seasonal_dt <- data.table(dte = prc_all[, dte],
                            crude = crude_seasonal)

fcst_list_crude_sea <- train_test_split_forecast_ts(y_dt = crude_seasonal_dt, test_points = 259)
forecast::autoplot(fcst_list_crude_sea[[2]])
```


```{r}
fcst_list_crude_reg <- train_test_split_forecast(y_dt = prc_all[, .(dte, crude)], test_points = 259, 
                                                  xreg_train = crude_seasonal_dt,
                                                  xreg_test = as.numeric(fcst_list_crude_sea[[1]]$mn)
                                                  )

plot_fcst_resid(y_dt = prc_all[, .(dte, crude)], fcst_dt = fcst_list_crude_reg[[1]], fcst = fcst_list_crude_reg[[2]], 
                train_dt = fcst_list_crude_reg[[3]], vis_start_date = lubridate::as_date("2021-07-01"))

```

### Adding Seasonality in Directly

The forecast still isn't producing an outcome we are hoping for. It is probably because a seasonality of 365 is too high for most models to be looking for. Therefore we are going to forecast the seasonality as before and then add it onto the crude forecast. This  seems to produce the better results.

```{r}
fcst_list_crude_reg <- train_test_split_forecast(y_dt = prc_all[, .(dte, crude)], test_points = 259)

fcst_dt_direct <- fcst_list_crude_reg[[1]][, mn := mn + as.numeric(fcst_list_crude_sea[[1]]$mn)]

plot_fcst_resid(y_dt = prc_all[, .(dte, crude)], fcst_dt = fcst_dt_direct, fcst = fcst_list_crude_reg[[2]], 
                train_dt = fcst_list_crude_reg[[3]], vis_start_date = lubridate::as_date("2021-01-01"))

```

## Ironore With Regressor

As mentioned earlier the regressor looks to be more predictive than the price itself, Therefore we are going to project the the comparison price and regress it onto the ironore to see if it gives a better prediction. 
 
```{r}

fcst_list_ironore <- train_test_split_forecast(y_dt = ironore, test_points = test_points)

fcst_list_ironore_reg <- train_test_split_forecast(y_dt = ironore, test_points = test_points, 
                                                  xreg_train = crude,
                                                  xreg_test = fcst_list_ironore[[1]]$mn
                                                  )

plot_fcst_resid(y_dt = ironore, fcst_dt = fcst_list_ironore_reg[[1]], fcst = fcst_list_ironore_reg[[2]], 
                train_dt = fcst_list_ironore_reg[[3]], vis_start_date = lubridate::as_date("2020-01-01"))

```

```{r}

row_ironore <- nrow(ironore)

ironore_diff365 <- data.table(dte = ironore[366:row_ironore, dte], ironore = diff(ironore[[2]], lag = 365))

fcst_list_ironore <- train_test_split_forecast(y_dt = ironore_diff365, test_points = test_points)

fcst_list_ironore_reg <- train_test_split_forecast(y_dt = ironore[(row_ironore-nrow(ironore_diff365)+1):row_ironore], test_points = test_points, 
                                                  xreg_train = ironore_diff365,
                                                  xreg_test = fcst_list_ironore[[1]]$mn
                                                  )

plot_fcst_resid(y_dt = ironore, fcst_dt = fcst_list_ironore_reg[[1]], fcst = fcst_list_ironore_reg[[2]], 
                train_dt = fcst_list_ironore_reg[[3]], vis_start_date = lubridate::as_date("2020-01-01"))

```

It looks like the models are struggling to find the patterns within the data so we are going to need to force it in slightly. Below is the forecast using the seasonally decomposed forecast of the comparison as a regressor. This means it wont have ans influence on the modelling process but the main aspects of the ironore forecast will take precedence. 

```{r}
row_ironore <- nrow(ironore)

ironore_seasonal <- decompose(ironore_ts)$seasonal

ironore_seasonal_dt <- data.table(dte = prc_all[, dte],
                            ironore = ironore_seasonal)

fcst_list_ironore_sea <- train_test_split_forecast_ts(y_dt = ironore_seasonal_dt, test_points = 259)
forecast::autoplot(fcst_list_ironore_sea[[2]])
```


```{r}
fcst_list_ironore_reg <- train_test_split_forecast(y_dt = prc_all[, .(dte, ironore)], test_points = 259, 
                                                  xreg_train = ironore_seasonal_dt,
                                                  xreg_test = as.numeric(fcst_list_ironore_sea[[1]]$mn)
                                                  )

plot_fcst_resid(y_dt = prc_all[, .(dte, ironore)], fcst_dt = fcst_list_ironore_reg[[1]], fcst = fcst_list_ironore_reg[[2]], 
                train_dt = fcst_list_ironore_reg[[3]], vis_start_date = lubridate::as_date("2021-07-01"))

```

### Adding Seasonality in Directly

The forecast still isn't producing an outcome we are hoping for. It is probably because a seasonality of 365 is too high for most models to be looking for. Therefore we are going to forecast the seasonality as before and then add it onto the ironore forecast. This  seems to produce the better results.

```{r}
fcst_list_ironore_reg <- train_test_split_forecast(y_dt = prc_all[, .(dte, ironore)], test_points = 259)

fcst_dt_direct <- fcst_list_ironore_reg[[1]][, mn := mn + as.numeric(fcst_list_ironore_sea[[1]]$mn)]

plot_fcst_resid(y_dt = prc_all[, .(dte, ironore)], fcst_dt = fcst_dt_direct, fcst = fcst_list_ironore_reg[[2]], 
                train_dt = fcst_list_ironore_reg[[3]], vis_start_date = lubridate::as_date("2021-01-01"))

```

